Browsernative Microservices

### 2017-01-27 paper.tex
pdf/tex build

Computers can scale, humans can't. Ever since, a program or complex
system made by humans has been constrained by our mental capabilities.
Like in the analogy of juggling balls, our brain can just ``juggle'' a
few things at a time. Rich Hickey, the inventor of the programming
language \emph{Clojure} gave an inspirational keynote on the topic of
\textbf{simplicity}.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=46s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} In every sphere
of a humans life, simplicity aligns perception with our mental
capacities. Understanding the word \textbf{simple} alone will outline
the this paper on its own.

Derived from the ancient Latin word \textbf{simplex}, simple can be
understood as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points directly to the unidimensional aspect. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
\emph{easy}, but while \emph{easy} is more of a relative nature, simple
can be laid out as a objective manner - a goal software development
should pursuit and praise.

@@

The \emph{state of the web} is certainly a different kind of complex
multidimensional beast. ``The web'' is coined to be meant ``everything
that runs in the browser''. While simplicity in the backend is mostly a
matter of principles, any frontend developer is highly restricted on the
entangled nature of the browser platform. Every call to an webpage
results in a single \textbf{global DOM tree} managed by the browser
engine. In the early days of the web a typical static page contained
just a bunch of text nodes and some styling. Nowadays a typical webapp
contains text nodes, functionality and styling from different resources
and is expected to work highly dynamic and frictionless.

In the last four years the average transfer size of a webpage doubled to
currently around 2.5 MB.\footnote{\href{http://httparchive.org/trends.php}{HTTPArchive
  Trends}} Subtracting images, fonts or other content the size of HTML,
CSS and JS sums up to a total average of 550 kb. One character weights
around 1 byte which means an average webpage is delivering 550.000
character or around 125 pages of single-spaced text. Frederic Filloux
calculated the ratio of real content on different newspaper websites and
came to the conclusion, that only round about 5-6 \% of the transferred
characters made for consumption.{[}1{]}

Having an 95 \% overhead is rather undesirable for both the consumers
and creators of the website. Since it's a widespread problem without a
single point of failure one can argue the platform itself is the
failure. In the recent years many frameworks, libraries and
methodologies tried to simplify the way of building user interfaces by
scoping assets and design rules into maintainable components. One of the
most adopted technique was the \textbf{virtual DOM} approach essential
ditching the old rule of separated HTML, CSS and JS in favor of
\textbf{scoped JS components} (containing HTML and CSS). While
frameworks are a valid approach to solve the problem they remain highly
opinionated.

In 2013 thinkers, creators and browser vendors joined together to
propose \emph{The Extensible Web Manifesto}.\footnote{\href{https://extensiblewebmanifesto.org/}{The
  Extensible Web Manifesto}} The claim of the Manifesto is to enhance
the current web with new low-level capacities. Those capacities should
empower creators of the web to write more declarative code and therefore
overcome known bottlenecks and artificial abstractions. Four years
later, the new capabilities of JavaScript leapfrogged and many new
low-level APIs brought to life. With the new APIs at hand a webdeveloper
can create scoped and highly reusable microservices without additional
libraries even directly in the browser console.

\emph{Disclaimer:} Currently, many of the new standards proposed
throughout this paper can only tried and tested inside Google Chrome or
Opera. As the innovation cycle is pretty short it is expected to see
those standards go live in all major browsers soon.In the meantime all
of those functionalities can be safely \textbf{polyfilled}.

\subsection{Microservices}\label{microservices}

Microserve

untangle the web

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

Betrachtet man andere populäre Webframeworks dieser Zeit, wie
beispielsweise React oder Angular, lässt sich ziemlich schnell ein
gemeinsames Designpattern ausmachen, wie Komponenten intern ihre
Zuständen verwalten. Der Facebook Entwickler Dan Abramov hat dieses
dichotome Pattern systematisch erfasst und in zwei Kategorien
eingeteilt.

Nach Abramov existieren zum einen Komponenten, die alleine für das
\textbf{Darstellen von Information} zuständig sind. Diese Komponenten
nehmen keinerlei Einfluss auf Informationen oder anderen Komponenten um
sich herum. Sie sind im wahrsten Sinne passiv und fremdgesteuert über
klar definierte Schnittstellen. Diese Komponenten sind häufig flexibel
einsetzbar und hochgradig wiederwerwendbar. Abramov nennt diese
Komponenten ``Presentational Components''.{[}2{]} Verortet man diese Art
von Komponenten innerhalb des \emph{MVC Pattern}, sind die Komponenten
reine \emph{View-Elemente}.

Im Gegensatz dazu stehen Komponenten, die für die \textbf{Verarbeitung
von Informationen} zuständig sind. Diese so genannten ``Container
Components'' haben oft einen internen Zustand, den sie verändern können
und an ihre Kindkomponenten weiterreichen können.{[}2{]} Im \emph{MVC
Pattern} handelt es sich um die \emph{Controller}. In der funktionalen
Programmiersprache Elm werden diese Elemente \emph{Updater} genannt, was
ihre Funktion noch besser umschreibt.

\section{Microservices}\label{microservices-1}

Opening up the case for \emph{Browsernative Microservices} brings up the
question about the concept microservices in general. In fact the concept
of microservices has many facets, stretching beyond disciplines and
technical boundaries. It lacks a formal standardization but there are
certain ideas emergine from this pattern. As a primary source of truth
this articles relies on the work of Sam Newman, who has written a
comprehensive guide in \emph{Building Microservices}. The purpose of
this section is to match those ideas against the manifestations of web
components.

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}3, p. 2{]} or with the words
of Fowler and Lewis: ``It is an approach to developing a single
application as a suite of small services, each running in its own
process and communicating with lightweight mechanisms,\ldots{}''{[}4{]}
Microservices incorporate many ideas, like \emph{domain-driven design}
where we try to represent the real world in our code.{[}3, p. 2{]} Or
making use of \emph{continuous delivery} for pushing software rapidly
through \emph{automated deployment} mechanisms in production.{[}4{]}
And, last but not least, microservices utilizes the idea of small teams
with a lot of product knowledge working mostly autonomous on their very
own service with their very own set of tools and techniques.

\subsection{Technical perspective}\label{technical-perspective}

@@

Talking about microservices in a browsernative context isn't that far
fetched as microservices themselves incorporate many ideas from the web.
Exemplary, microservices often communicate via an HTTP request-response
with resource API's and lightweight messaging.{[}4{]} Nevertheless,
while (server-side) microservices offer wide ranges of technical
possibilities, we must take into account that (client-side) browsers
come with certain constraints and limitations.

Fowler and Lewis issue a call for using services as components. A
component~is regarded as a unit of software that is independently
replaceable and upgradeable. The main advantage of a component in
contrast of library is the possibility of an independent deployment. It
aligns perfectly with the main goal of a microservice architecture to
strip away most of the dependencies in favour of clean
interfaces.{[}4{]}

Components

First of all, from a technical perspective, a microservice reinforces
the \emph{Single Responsibility Principle} defined by Robert C. Martin:
``Gather together those things that change for the same reason and
separate those things that change for different reasons.''{[}5{]} An a
way this principle tackles another often cited design principle of the
\emph{seperation of concerns}. Web Components incorporate this principle
in multiple ways while still remaining flexible.

@@

The concept of microservices incooperates not only a technical
perspective. Microservice patterns are a product of real-world
usage.{[}3, p. 1{]} In a real world we typical have to deal with the so
called \emph{Conway's Law}:

@@

Like \emph{Custom Events}, \emph{Web Workers} had been around for a long
time and therefore embrace full support among major browser vendors.
They emerged at around 2009 when discussions about browser performance
was in the early stages of development. Nevertheless, the addressed
problem of \emph{Web Workers} is a fundamental language problem of JS
itself.

@@

\hypertarget{ref-Abramov2015}{}
{[}2{]} D. Abramov, ``Presentational and Container Components --
Medium.'' 2015 {[}Online{]}. Available:
\url{https://medium.com/@dan_abramov/smart-and-dumb-components-7ca2f9a7c7d0}.
{[}Accessed: 01-Dec-2016{]}

>>>>

Computers can scale, humans can't. Ever since a program or complex
system made by humans has been constrained by humans mental
capabilities. Like in the analogy of juggling balls, our brain can just
``juggle'' a few things at a time. Rich Hickey, the inventor of the
programming language \emph{Clojure} gave an inspirational keynote on the
topic of \textbf{simplicity}.\footnote{\href{https://www.youtube.com/watch?v=rI8tNMsozo0\&t=46s}{Rails
  Conf 2012 Keynote: Simplicity Matters by Rich Hickey}} In every sphere
of a humans life, simplicity aligns perception with our mental
capacities. This paper should be read under the impression of
simplicity.

Derived from the ancient Latin word \textbf{simplex}, simple can be
understood as ``literally, uncompounded or onefold''\footnote{\href{http://www.etymonline.com/index.php?term=simple}{Etymology
  Dictionary}} which points directly to the unidimensional aspect. While
complexity describes the multilayered und entangled nature of
conditions, simplicity empowers the human brain to reason about issues
in a straightforward manner. It certainly has some overlapping's with
easy, but while easy is more of a relative nature, simple can be laid
out as a objective manner - a goal software development should pursuit
and praise.

@@

The \textbf{state of the web} is certainly a different kind of
complexity beast. ``The web'' is coined to ``everything that runs in the
browser''. While simplicity in the backend is mostly a matter of
principles, any frontend developer is highly restricted on the highly
deterministic nature of the browser platform.

In the last four years the average transfer size of a webpage doubled to
currently around 2.5 MB.\footnote{\href{http://httparchive.org/trends.php}{HTTPArchive
  Trends}} Subtracting images, fonts or other content the size of HTML,
CSS and JS sums up to a total average of 550 kb. One character weights
around 1 byte which means an average webpage is delivering 550.000
character or around 125 pages of single-spaced text. Frederic Filloux
calculated the ratio of real content on different newspaper websites and
came to the conclusion, that only round about 5-6 \% of the transferred
characters made for human consumption.{[}1{]}

Having an 95 \% overhead is rather undesirable for both the consumers
and creators of the website. Since it's a widespread problem without a
single point of failure one can argue the platform itself is the
failure. By design, every pageload results in a monolithic DOM tree
managed by the browser engine. Whether rendering just a bunch of static
text nodes or an ever changing webapp the underlying global nature of
the DOM tree remains the same. Every additional service added to the
webapp embodies another fold of complexity.

In an non-deterministic runtime environment, encapsulation and
modularization is a typical pattern to make complexity manageable and to
accommodate future uncertainty.{[}2, p. 1{]} Today many websites tend
towards becoming complex single-page applications (and therefore the
average JS payload is soaring too). Consequently, the demands to the
browser platform changed from a static page renderer to a
\textbf{complexe UI machine}. Under the current situation only
additional layers of abstraction can handle those complexity.

In the recent years many \textbf{frameworks}, libraries and
methodologies tried to tackle the global nature of the DOM by scoping
assets and design rules into maintainable components. While the DOM
can't be scoped, JS can. Therefore many of those frameworks, like
ReactJS, AngularJS or VueJS just to name a few, ditched the old rule of
separated HTML, CSS and JS in favor of an additional layer of abstracted
JS components (containing content, markup and styling). Quiet often
those frameworks mimic a MVC pattern on top of the browser engine which
is a reasonable simple design pattern to build graphical user
interfaces. This paper will apply the same pattern for building a
browsernative microservice.While frameworks are a valid approach to
build scalable web applications they remain highly opinionated, embody
inherent complexity themselves and can change and break over time.
Furthermore, they inflate the code with libraries. A standardized way
for creating complex UI interfaces painlessly requires new build-in
browser capabilities.

In 2013 thinkers, creators and browser vendors joined together to
propose \emph{The Extensible Web Manifesto}.\footnote{\href{https://extensiblewebmanifesto.org/}{The
  Extensible Web Manifesto}} The claim of the manifesto is to enhance
the current web platforms with new low-level capabilities. Those
features should empower creators of the web to write more declarative
code and therefore overcome known bottlenecks and artificial
abstractions. Four years later, the enhancement of JavaScript
leapfrogged and many new low-level APIs brought to life. With the new
APIs at hand a webdeveloper can create scoped and highly reusable
microservices without additional libraries even directly in the browser
console.

\emph{Disclaimer:} Currently, all new technologies proposed throughout
this paper can be tried and tested in Google Chrome, Opera and the
Safari Technology Preview. Usually the browser adoption time is short in
time an therefore full support of all standards is expected soon.
Meanwhile this standards can be used through \textbf{polyfills} on every
important legacy browser.

\section{Microservices}\label{microservices}

Calling the case for \emph{Browsernative Microservices} brings up the
question about the concept of microservices in general. In fact the
concept of microservices has many facets, stretching beyond disciplines
and technical boundaries. It lacks a formal standardization but subsumes
certain ideas emergine from this pattern. As a primary source of truth
this paper relies on the work of Sam Newman, who has written a
comprehensive guide called \emph{Building Microservices} and the work of
Fowler and Lewis. The purpose of this section is to match those ideas
against the browser platform as targeted runtime.

In a nutshell a microservice is a small, autonomous service that works
together with other services seamlessly.{[}3, p. 2{]} or with the words
of Fowler and Lewis: ``\ldots{} the microservice architectural style is
an approach to developing a single application as a suite of small
services, each running in its own process and communicating with
lightweight mechanisms, often an HTTP resource API''{[}4{]}
Microservices incorporate many other ideas, like \emph{domain-driven
design} where we try to represent the real world in our code.{[}3, p.
2{]} Or making use of \emph{continuous delivery} for pushing software
rapidly through \emph{automated deployment} mechanisms in
production.{[}4{]} And, last but not least, microservices utilizes the
idea of small teams with a lot of product knowledge working mostly
autonomous on their very own service with their very own set of tools
and techniques.

\subsection{Componentization via
Services}\label{componentization-via-services}

``A~\textbf{component}~is a unit of software that is independently
replaceable and upgradeable.''{[}4{]} Compared to libraries, a component
service offers multiple advantages. A library is only loosely coupled to
the implementation and therefore hard to track in functionality.
Changing a library often requires an unforeseen amount of time fixing
implementations. Therefore in the field we often see multiple versions
of libraries deployed simultaneous (like with JQuery) which adds a huge
amount of overhead on the page load. Eventually, dead code elimination
won't work in highly coupled libraries. Components instead incorporate a
much better evolution mechanism in the service contracts. Changing
functionality won't break other components. A component can
progressively enhanced which guarantees functionality throughout
different versions.

Compared to libraries componentized services offers a more explicit
interface.{[}4{]} While the functionality of a library needs
documentation, a component functionality is exposed via the components'
signature and therefore highly accessible. HTML elements are explicitly
controlled via their attributes and the new HTML service component are
no exception to that.

A microservice reinforces the \emph{Single Responsibility Principle}
defined by Robert C. Martin: ``Gather together those things that change
for the same reason and separate those things that change for different
reasons.''{[}5{]}

\subsection{Organized around Business
Capabilities}\label{organized-around-business-capabilities}

Microservice patterns are a product of real-world usage.{[}3, p. 1{]}
Instead of splitting teams along the technology stack (UI Experts
-\textgreater{} Middleware -\textgreater{} Database) a microservice
approach model teams around \textbf{business capabilities}.{[}4{]}
Consequently every team is capable of planning, designing, implementing,
testing and maintaining their very own microservice. Across the stack
every member gains high competence about the functionality of the
service.

By assigning service responsibility a team, the so called
\textbf{Definition of Done} (DoD) shifts from ``accomplishing projects''
to ``accomplishing products''. This new paradigm not only changes the
administrative overhead like budgeting or resource allocation. It
creates a kind of responsibility connection from the team to the
service. Expectedly, those teams are more motivated within their very
own service and exhibit a more sophisticated iteration time.{[}4{]}

For many companies working in the spheres of the internet, the
client-side is the most crucial area of business success. Todays
client-side includes many devices of differing sizes and
functionalities. As browser technologies, design guidelines and devices
change in a fast paced manner it makes absolutely sense to shift
responsibility towards the teams altogether. A \emph{browsernative
microservice} needs commitment over its whole lifecycle.

\subsection{Smart endpoints and dumb
pipes}\label{smart-endpoints-and-dumb-pipes}

Probably most microservices build for the browser are impure in
functionality. They aim for side-effects like changing the UI given to
input parameters or emitting browser events. Therefore a web developer
must design communication channels for those microservices keeping in
mind the nested nature of the DOM.

\emph{Smart endpoints and dumb pipes} is coined to the approach of
designing communication mostly decoupled and as cohesive as possible.
This essentially removes most of the error-prone middleware altogether.
In the analogy of the real world a message channel should look like
sending a letter: Only two endpoints (sender and receiver) connoted as
``smart'' and the letter itself looks mostly dumb (but unified) from the
outside. Only the receiver is able to process the content message.

Cited earlier in this paper microservices often rely on simple HTTP
request-response with resource API's and lightweight messaging as
communication protocols.{[}4{]} In the later parts of this paper the
reader will see a generic \emph{browsernative microservice} split in the
fashion of \textbf{MVC}. In the \textbf{VIEW} the microservice will rely
on browser events and in-memory calls while the \textbf{CONTROLLER} have
to deal with other resources over the fetch API.

\subsection{Decentralized Governance}\label{decentralized-governance}

\subsection{Decentralized Data
Management}\label{decentralized-data-management}

\subsection{Infrastructure Automation}\label{infrastructure-automation}

\subsection{Design for failure}\label{design-for-failure}

\subsection{Evolutionary Design}\label{evolutionary-design}

@@

@@

The concept of microservices incooperates not only a technical
perspective.

@@

Like \emph{Custom Events}, \emph{Web Workers} had been around for a long
time and therefore embrace full support among major browsers. They
emerged at around 2009 when discussions about browser performance was
still in the early days. Nevertheless, the addressed problem of
\emph{Web Workers} is a fundamental language problem of JS itself.

@@

\hypertarget{ref-Baldwin2006}{}
{[}2{]} C. Y. Baldwin and K. B. Clark, ``Modularity in the Design of
Complex Engineering Systems,'' in \emph{Complex engineered systems:
Science meets technology}, D. Braha, A. A. Minai, and Y. Bar-Yam, Eds.
Berlin, Heidelberg: Springer Berlin Heidelberg, 2006, pp. 175--205
{[}Online{]}. Available:
\href{http://dx.doi.org/10.1007/3-540-32834-3\%7B/_\%7D9}{http://dx.doi.org/10.1007/3-540-32834-3\{\textbackslash{}\_\}9}
